## Linear regression models for Kumpula station 
This section is divided into (1) model prediction and (2) analysis and results.
We predict the linear regression model for the Kumpula station for the four seasons and for the 64 forecast
periods. For ease of analysis we use only the forecasts that are at 00:00 UTC (analysis_time is 1). We use the simple linear regression(lm), linear regression with step method and linear regression with cross validation (glm.cv).

The linear regression model takes two response variables namely T2Obs and D2Obs, the observations of air temperature and dewpoint temperature at 2 meters. As the predictors we use the ECMWF model outputs. All these variables are explained in the Section on weather data. Using the linear regression models, mymodel, lm, lmstep and glmcv, we predict the
T2 and D2, the air temperature and dewpoint temperature for 64 forecast periods. We then calculate the relative humidity, RH, using the predicted T2 and D2 using the four  linear regression models. We also calculate the RH using the T2 and D2 from the ECMWF raw model. The relative humidity is calculated using some standard formulas.

The R program is done in a modularised way with functions.

### Linear model prediction
```{r, warning=FALSE, message=F, echo=TRUE} 

# Packages 
# reshape package for melt operation 
library(reshape2)

# glm and glm.cv
library(boot)

FitWithmymodel <- function(training.set, test.set = NULL, parameter) {
  # Estimates a  linear model using "mymodel" we developed and predicts with the model.
  #
  # Args:
  #   training.set:
  #   test.set:
  #   parameter shows whether we want to model for T2 or D2.
  #   introducing the parameter is a simple way to select the formula that is to be 
  #   used for T2 and D2
  #
  # Returns:
  #   A list of coefficients, residuals and fitted values.
  
  response <- colnames(training.set)[1]
  if (parameter == 1) { 
    model <- lm(T2Obs ~ T2 + D2 + T_950 + T_925 + T2_M1 + D2_M1 + SKT + LCC, data = training.set )
  }
  else if (parameter == 2) {
   model <- lm(D2Obs ~ T2 + D2 + T_950 + T_950_M1 + D2_M1 + TP + RH_950 + V10 + LCC, data = training.set)
  }
  
  
  if (is.null(test.set)) {
    test.set <- training.set
  }
  
  fitted.values <- predict(object = model, newdata = test.set)
  residuals <- test.set[,1] - fitted.values
  
  fitted.values <- as.numeric(fitted.values)
  residuals <- as.numeric(residuals)
  
  results <- list("coefficients" = model$coefficients,
                  "residuals" = residuals,
                  "fitted.values" = fitted.values)
  
  class(results) <- "simplelm"
  
  return(results)
}

FitWithLM <- function(training.set, test.set = NULL) {
  # Estimates a full linear model and predicts with the model.
  #
  # Args:
  #   training.set:
  #   test.set:
  #
  # Returns:
  #   A list of coefficients, residuals and fitted values.
  
  response <- colnames(training.set)[1]
  formula <- paste(response, " ~ .", sep = "")
  model <- lm(formula = formula, data = training.set)
  
  if (is.null(test.set)) {
    test.set <- training.set
  }
  
  fitted.values <- predict(object = model, newdata = test.set)
  residuals <- test.set[,1] - fitted.values
  
  fitted.values <- as.numeric(fitted.values)
  residuals <- as.numeric(residuals)
  
  results <- list("coefficients" = model$coefficients,
                  "residuals" = residuals,
                  "fitted.values" = fitted.values)
  
  class(results) <- "simplelm"
  
  return(results)
}

FitWithStep <- function(training.set, test.set = NULL,
                        object.formula, upper.formula, lower.formula,
                        direction, steps = 1000) {
  
  
  # Chooses a linear model by a stepwise algorithm and predicts a fit on an independant test data
  #
  # Args:
  #   training.set: a data matrix
  #   test.set: a data matrix
  #   object.formula: a formula viable for a linear model
  #   upper.formula: a formula viable for a linear model
  #   lower.formula: a formula viable for a linear model
  #   direction: both/forward/backward
  #   steps: the number of steps the algorithm is allowed to take
  # 
  # Returns:
  #   A list of coefficients, residuals and fitted values.
  
  # saving the training and the test set to global environment
  training.set <- training.set
  test.set <- test.set
  
  # generating the linear models
  object.lm <- lm(object.formula, data = training.set)
  upper.lm <- lm(upper.formula, data = training.set)
  lower.lm <- lm(lower.formula, data = training.set)
  
  # choosing the best model
  step.model <- step(object = object.lm,
                     scope = list(upper = upper.lm, lower = lower.lm),
                     direction = direction, trace = 0 , steps = steps)
  
  # if there are no test data, the fit is calculated using the training data
  if (is.null(test.set)) {
    test.set <- training.set
  }
  
  # calculating the fit
  fitted.values <- predict(object = step.model, newdata <- test.set)
  
  fitted.values <- as.numeric(fitted.values)
  residuals <- as.numeric(test.set[,1] - fitted.values)
  coefficients <- step.model$coefficients
  
  results <- list("coefficients" = coefficients,
                  "residuals" = residuals,
                  "fitted.values" = fitted.values)
  
  class(results) <- "simplelm"
  return(results)
}


FitWithglmcv<- function(training.set, test.set = NULL, k) {
  # Estimates a full glm linear model with cross validation  and predicts with the model.
  #
  # Args:
  #   training.set:
  #   test.set:
  #   number of folds
  # Returns:
  #   A list of coefficients, residuals and fitted values.
  
  response <- colnames(training.set)[1]
  formula <- paste(response, " ~ .", sep = "")
  model <- lm(formula = formula, data = training.set)
  
  set.seed(123)
  cv.error.10 = rep(0, 10)
  for (i in 1:10) {
    model  = glm(formula = formula, data = training.set)
    cv.error.10[i] = cv.glm(training.set, model, K = k)$delta[1]
    cv.error.10
}

    
  if (is.null(test.set)) {
    test.set <- training.set
  }
  
  fitted.values <- predict(object = model, newdata = test.set)
  residuals <- test.set[,1] - fitted.values
  
  fitted.values <- as.numeric(fitted.values)
  residuals <- as.numeric(residuals)
  
  results <- list("coefficients" = model$coefficients,
                  "residuals" = residuals,
                  "fitted.values" = fitted.values)
  
  class(results) <- "simplelm"
  
  return(results)
}


ModelsOfAlgorithms <- function(training.set, test.set, parameter) {
  # Models generated using different algorithms, currently with lm, lmstep and glmcv
  # 
  #
  # Args:
  #   training.set:
  #   test.set:
  #
  # Returns:
  # A list of elements ' fitted_models', 'residuals', and 'response' variable
  
  # model details 
  response <- colnames(training.set)[1]
  fInitial <- as.formula(paste(response, " ~ 1", sep = ""))
  fFull <- as.formula(paste(response, " ~ .", sep = ""))
  
  # Create a list to store the model results for T2 and D2 observations
  fitted.values.by.algorithm <- as.data.frame(matrix(NA, ncol = 0, 
                                                        nrow=nrow(test.set)))
  residuals.by.algorithm <- as.data.frame(matrix(NA, ncol = 0,
                                                    nrow = nrow(test.set)))
 
  elapsed.time.by.algorithm <- NULL
 
   # mymodel
    tmp.time.result <- system.time(
    tmp.result <- FitWithmymodel(
                    training.set, test.set, parameter
                  )
  )
  
  fitted.values.by.algorithm$mymodel <- tmp.result$fitted.values
  residuals.by.algorithm$mymodel  <- tmp.result$residuals
  elapsed.time.by.algorithm <- c(elapsed.time.by.algorithm,
                                mymodel = tmp.time.result[[3]])
  
  
  
  # lm 
  tmp.time.result <- system.time(
    tmp.result <- FitWithLM(
                    training.set, test.set
                  )
  )
  
  fitted.values.by.algorithm$lm <- tmp.result$fitted.values
  residuals.by.algorithm$lm <- tmp.result$residuals
  elapsed.time.by.algorithm <- c(elapsed.time.by.algorithm,
                                lm = tmp.time.result[[3]])
  
  
  
  # lm step model 
  
  tmp.time.result <- system.time(
    tmp.result <- FitWithStep(
                    training.set, 
                    test.set, 
                    fInitial, fFull, fInitial, 
                    direction = "forward", steps = 1000)
  )
  
  fitted.values.by.algorithm$lmstep <- tmp.result$fitted.values
  residuals.by.algorithm$lmstep <- tmp.result$residuals
  elapsed.time.by.algorithm<- c(elapsed.time.by.algorithm,
                                    lmstep =tmp.time.result[[3]])
  
 
  # glm cv model
   
  tmp.time.result <- system.time(
    tmp.result <- FitWithglmcv(
      training.set, 
      test.set, 
      k = 4
      )
  )
  
  fitted.values.by.algorithm$glmcv <- tmp.result$fitted.values
  residuals.by.algorithm$glmcv <- tmp.result$residuals
  elapsed.time.by.algorithm <- c(elapsed.time.by.algorithm,
                                 glmcv = tmp.time.result[[3]])
  
   results  <-  list(fitted.values = fitted.values.by.algorithm,
                      residuals = residuals.by.algorithm,
                      elapsed.time = elapsed.time.by.algorithm)
  
  
  return(results)
}


  
CalculateRH <- function(wdata) {
  # Calculates the realative humidity RH using various fitting algorithms
  #
  # Args
  #   df: dataframe 
  #
  # Returns:
  #   rmse.pred.by.model: rmse of lmstep and glmcv predictions
  #   rmse.RH.by.model: rmse RH values from the ecmwf, lm, lmstep and glmcv
  
   # Create a list to store the rmse values for predictions and RH using the models values, 
   # predicted values using lm, lmstep, glmcv
  rmse.pred.by.model <- NULL
  rmse.RH.by.model <- NULL
 
  # Separating T2 and D2 and generating T2_Obs and modeldata aswellas D2_Obs and modeldata
  
  T2_Obs_model_data <- wdata[, c(6:(ncol(wdata)-1))]
  D2_Obs_model_data <- cbind(wdata[ncol(wdata)], wdata[,7:(ncol(wdata)-1)])

  
  
   # Splitting into train and test data
  n <- nrow(T2_Obs_model_data)
  train = sample(1:n, size = round(0.75*n), replace=FALSE)
  T2_Obs_model_data_train <- T2_Obs_model_data[train,]
  T2_Obs_model_data_test <-  T2_Obs_model_data[-train,]
  D2_Obs_model_data_train <- D2_Obs_model_data[train,]
  D2_Obs_model_data_test <-  D2_Obs_model_data[-train,]
  
   
  # Models and predictions for T2 and D2. Parameter value selects the formula for mymodel
  
  results.T2 <- ModelsOfAlgorithms(T2_Obs_model_data_train, T2_Obs_model_data_test, parameter = 1)
  results.D2 <- ModelsOfAlgorithms(D2_Obs_model_data_train, D2_Obs_model_data_test, parameter = 2)
  
  
  # T2 and D2 observation and model data  
  T2_obs_test <- T2_Obs_model_data_test[,1]
  D2_obs_test <- D2_Obs_model_data_test[,1]
  
  T2_model_test <- T2_Obs_model_data_test[,3]
  D2_model_test <- D2_Obs_model_data_test[,4]
  
  
  # T2 and D2 from predictions using mymodel, lm, lmstep and glmcv
  
  
  T2_pred_mymodel <- results.T2$fitted.values$mymodel
  T2_pred_lm <- results.T2$fitted.values$lm
  T2_pred_lmstep <- results.T2$fitted.values$lmstep
  T2_pred_glmcv<- results.T2$fitted.values$glmcv

  D2_pred_mymodel <- results.D2$fitted.values$mymodel
  D2_pred_lm <- results.D2$fitted.values$lms
  D2_pred_lmstep <- results.D2$fitted.values$lmstep
  D2_pred_glmcv<- results.D2$fitted.values$glmcv

  # rmse for T2 and  D2 obtained using the rawmodel, mymodel, lm, lmstep and glmcv  models
  
  rmse.T2.raw <- sqrt(mean((T2_obs_test - T2_model_test)^2))
  rmse.T2.mymodel <- sqrt(mean(results.T2$residuals$mymodel^2))
  rmse.T2.lm <- sqrt(mean(results.T2$residuals$lm^2))
  rmse.T2.lm <- sqrt(mean(results.T2$residuals$lm^2))
  rmse.T2.lmstep <- sqrt(mean(results.T2$residuals$lmstep^2))
  rmse.T2.glmcv <- sqrt(mean(results.T2$residuals$glmcv^2))
 
  rmse.D2.raw <- sqrt(mean((D2_obs_test - D2_model_test)^2))
  rmse.D2.mymodel <- sqrt(mean(results.D2$residuals$mymodel^2))
  rmse.D2.lm <- sqrt(mean(results.D2$residuals$lm^2))
  rmse.D2.lmstep <- sqrt(mean(results.D2$residuals$lmstep^2))
  rmse.D2.glmcv <- sqrt(mean(results.D2$residuals$glmcv^2))

   rmse.pred.by.model<- c(rmse.pred.by.model,
                        T2raw = rmse.T2.raw,
                        T2mymodel = rmse.T2.mymodel,
                        T2lm = rmse.T2.lm,
                        T2lmstep = rmse.T2.lmstep,
                        T2glmcv = rmse.T2.glmcv,
                        D2raw = rmse.D2.raw,
                        D2mymodel = rmse.D2.mymodel,
                        D2lm = rmse.D2.lm,
                        D2lmstep = rmse.D2.lmstep,
                        D2glmcv = rmse.D2.glmcv
                    )
   
   
   
  # Relative humidity (RH) calculation
   
  saturation_humidity_obs <-  6.112*exp((17.62*T2_obs_test)/(T2_obs_test + 243.12))
  specific_humidity_obs <-  6.112*exp((17.62*D2_obs_test)/(D2_obs_test + 243.12))
  RH_obs <- 100*(specific_humidity_obs/saturation_humidity_obs)
 
  saturation_humidity_model <-  6.112*exp((17.62*T2_model_test)/(T2_obs_test + 243.12))
  specific_humidity_model <-  6.112*exp((17.62*D2_model_test)/(D2_obs_test + 243.12))
  RH_model <- 100*(specific_humidity_model/saturation_humidity_model)
  
  # predicted T2, D2 using mymodel, lm, lmstep and glmcv
  
  saturation_humidity_pred_mymodel <-  6.112*exp((17.62*T2_pred_mymodel)/(T2_pred_mymodel + 243.12))
  specific_humidity_pred_mymodel <-  6.112*exp((17.62*D2_pred_mymodel)/(D2_pred_mymodel + 243.12))
  RH_pred_mymodel <- 100*(specific_humidity_pred_mymodel/saturation_humidity_pred_mymodel)
  
  saturation_humidity_pred_lm <-  6.112*exp((17.62*T2_pred_lm)/(T2_pred_lm + 243.12))
  specific_humidity_pred_lm <-  6.112*exp((17.62*D2_pred_lm)/(D2_pred_lm + 243.12))
  RH_pred_lm <- 100*(specific_humidity_pred_lm/saturation_humidity_pred_lm)
  
  saturation_humidity_pred_lmstep <-  6.112*exp((17.62*T2_pred_lmstep)/(T2_pred_lmstep + 243.12))
  specific_humidity_pred_lmstep <-  6.112*exp((17.62*D2_pred_lmstep)/(D2_pred_lmstep + 243.12))
  RH_pred_lmstep <- 100*(specific_humidity_pred_lmstep/saturation_humidity_pred_lmstep)
  
  
  saturation_humidity_pred_glmcv <-  6.112*exp((17.62*T2_pred_glmcv)/(T2_pred_glmcv + 243.12))
  specific_humidity_pred_glmcv <-  6.112*exp((17.62*D2_pred_glmcv)/(D2_pred_glmcv + 243.12))
  RH_pred_glmcv <- 100*(specific_humidity_pred_glmcv/saturation_humidity_pred_glmcv)
  
  
  # rmse for RH
  rmse_RH_model <- sqrt(mean((RH_model - RH_obs)^2))
  rmse_RH_mymodel <- sqrt(mean((RH_pred_mymodel - RH_obs)^2))
  rmse_RH_lm <- sqrt(mean((RH_pred_lm - RH_obs)^2))
  rmse_RH_lmstep <- sqrt(mean((RH_pred_lmstep - RH_obs)^2))
  rmse_RH_glmcv <- sqrt(mean((RH_pred_glmcv - RH_obs)^2))
 
  
  rmse.RH.by.model<- c(rmse.RH.by.model,
                       rawmodelRH = rmse_RH_model,
                       mymodelRH = rmse_RH_mymodel,
                       lmRH = rmse_RH_lm,
                       lmstepRH = rmse_RH_lmstep,
                       glmcvRH = rmse_RH_glmcv
   )
  
  
  #cat("RH model rmse=", rmse_RH_model, "RH lm rmse=", rmse_RH_lm, 
  #     "RH lmstep rmse=", rmse_RH_lmstep,  "RH glmcv rmse=", rmse_RH_glmcv)
 
  results <- list(rmse.pred.by.model = rmse.pred.by.model,
                  rmse.RH.by.model = rmse.RH.by.model)
 return(results)
  
  
}


### The Main Program 
# Find the lm, lmstep and glmcv, linear regression models for  Kumpula station data. 
# We predict the T2 and D2, the air temperature and dewpoint temperature at 2 meters 
# using the above models. We calculate the relative humidy (RH) with the predcited T2 and D2.
# We compare T2, D2 and RH of the  ECMWF model with the predcited T2, D2 and RH.


# Initializing variables and data structures
station_id <- 2998
analysis_time <- 1 


rmse.T2D2 <- NULL
rmse.RH <- NULL
stn.season.atime.fperiod <- NULL
rmse.pred.station.season.atime.fperiod  <- NULL
rmse.rh.station.season.atime.fperiod  <- NULL

# loading the Kumpula data from the file data/station2998_T2_D2_Obs_model_data_with_timestamps.csv.
# This data is created using our program "create_weather_data.R"

T2_D2_Obs_model_data_with_timestamps = read.table(file = "data/station2998_T2_D2_Obs_model_data_with_timestamps.csv", sep = ",", header = TRUE)


# We separately model the data for the four seasons and 64 forecast periods (2:64)

for (s  in 1:4) {
  for(f in 2:65) { 

    df <- T2_D2_Obs_model_data_with_timestamps[which(T2_D2_Obs_model_data_with_timestamps$season == s &
                                                      T2_D2_Obs_model_data_with_timestamps$analysis_time == 1 & 
                                                           T2_D2_Obs_model_data_with_timestamps$forecast_period == f ),]
    
# The function CalculateRH, calculate the relative humidity from the predicted T2 and D2 using th elinear models.
    
    results <- CalculateRH(df)
    rmse.T2D2 <- rbind(rmse.T2D2, results$rmse.pred.by.model)
    rmse.RH <- rbind(rmse.RH, results$rmse.RH.by.model)
              
    stn.season.atime.fperiod <-  c(station = station_id,
                                   season = s,
                                   atime = analysis_time,
                                   fperiod = f
                                   )
     
      
      
     
      rmse.pred.station.season.atime.fperiod <- rbind(rmse.pred.station.season.atime.fperiod, c(stn.season.atime.fperiod, results$rmse.pred.by.model))                 
      rmse.rh.station.season.atime.fperiod <- rbind(rmse.rh.station.season.atime.fperiod, c(stn.season.atime.fperiod, results$rmse.RH.by.model))
              
    
  }
}



```
### Results

####Quartiles of T2 and D2

```{r, warning=FALSE, message=F, echo=TRUE} 

df.rmse.pred <- as.data.frame(rmse.pred.station.season.atime.fperiod)




for (s in 1:4) { 
    dfname <- paste("df_",s,sep ="")
    assign(dfname, df.rmse.pred[which(df.rmse.pred$season == s),])
}

# Winter 
summary(df_1[,5:14])

# Spring
summary(df_2[,5:14])

# Summer
summary(df_3[,5:14])

# Autumn
summary(df_4[,5:14])
```


```{r, warning=FALSE, message=F, echo=TRUE} 
df.rmse.rh <-  as.data.frame(rmse.rh.station.season.atime.fperiod)
for (s in 1:4) { 
    dfname <- paste("df_",s,sep ="")
    assign(dfname, df.rmse.rh[which(df.rmse.rh$season == s),])
}

# Winter 
summary(df_1[,5:9])

# Spring
summary(df_2[,5:9])

# Summer
summary(df_3[,5:9])

# Autumn
summary(df_4[,5:9])


```



### Plots
```{r, warning=FALSE, message=F, echo=TRUE} 

require(ggplot2)
require(reshape)

df.rmse.pred <- as.data.frame(rmse.pred.station.season.atime.fperiod)
df.rmse.rh <-  as.data.frame(rmse.rh.station.season.atime.fperiod)


df_winter_pred_model <- df.rmse.pred[which(df.rmse.pred$season == 1),]
df_winter_rh_model <- df.rmse.rh[which(df.rmse.rh$season == 1),]

df_winter_T2_pred_model <- df_winter_pred_model[, c(4:9)]
df_winter_D2_pred_model <- df_winter_pred_model[, c(4,10:14)]
fperiod <- df_winter_T2_pred_model$fperiod


molten <- melt(df_winter_T2_pred_model, id.vars = 'fperiod', variable_name = "model")
ggplot(molten, aes(x=fperiod, y=value, colour = model )) + geom_line()+xlab("forecast period") + ylab("RMSE") + ggtitle("RMSE in T2 prediction with  linear regression models for the winter season")

molten <- melt(df_winter_D2_pred_model, id.vars = 'fperiod', variable_name = "model")
ggplot(molten, aes(x=fperiod, y=value, colour = model )) + geom_line()+xlab("forecast period") + ylab("RMSE") + ggtitle("RMSE in D2 prediction with  linear regression models for the winter season")

df_winter_rh_model <- df_winter_rh_model[, c(4:9)]

molten <- melt(df_winter_rh_model, id.vars = 'fperiod', variable_name = "model")
ggplot(molten, aes(x=fperiod, y=value, colour = model )) + geom_line()+xlab("forecast period") + ylab("RMSE") + ggtitle("RMSE in RH with  linear regression models for the winter season")
```

### RMSE T2 in all seasons
```{r, warning=FALSE, message=F, echo=TRUE} 

df.rmse.pred.station.fperiod <- df.rmse.pred[,c(2, 4:9)]

molten <- melt(df.rmse.pred.station.fperiod , id.vars = c("season", "fperiod"), variable_name = "model")
ggplot(molten, aes(x= season, y=value, colour = model )) + geom_boxplot()+xlab("seasons(1:4)") + ylab("RMSE") + ggtitle("RMSE in T2 with  linear regression models for four seasons")
```


### RMSE D2 in all seasons
```{r, warning=FALSE, message=F, echo=TRUE} 

df.rmse.pred.station.fperiod <- df.rmse.pred[,c(2, 4,10:14)]

molten <- melt(df.rmse.pred.station.fperiod , id.vars = c("season", "fperiod"), variable_name = "model")
ggplot(molten, aes(x= season, y=value, colour = model )) + geom_boxplot()+xlab("seasons(1:4)") + ylab("RMSE") + ggtitle("RMSE in D2 with  linear regression models for four seasons")
```

### RMSE in all seasons
```{r, warning=FALSE, message=F, echo=TRUE} 

df.rmse.rh.station.fperiod <- df.rmse.rh[,c(2, 4:9)]

molten <- melt(df.rmse.rh.station.fperiod , id.vars = c("season", "fperiod"), variable_name = "model")
ggplot(molten, aes(x= season, y=value, colour = model )) + geom_boxplot()+xlab("seasons(1:4)") + ylab("RMSE") + ggtitle("RMSE in RH with  linear regression models for four seasons")
```


## Disussions

## Conclusions and future directions

## Ackmowledgements

I am grateful to the POSSE project manager Mikko Rauhala for allowing me to present the work I have been doing in the project as my final assignment for the IODS course. I am grateful to Jussi Ylhäisi for giving the problem of finding the interdependence of variables, providing me with the data on Ku,pula weather station and for the everyday  discussions on weather data and modelling. My deep gratitude goes  to the IODS course instructors, my fellow students whose assignments I have peer reviewed  and  who have reviewed my course work. They have taught me how to understand a data science problem, comprehend the variables in the problem and analyze the problem deeply. 

